### 2015-07-22 Harris_2015/manuscript-materials/draft.Rmd
draft

While competition doesn't reliably leave the expected pattern in community
structure (a low co-occurrence rate at the landscape level), it nevertheless does
leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
there will be a deficit of co-occurrences between the competing shrub species,
and *among unshaded sites*, there will also be such a deficit. More generally,
we can obtain much better estimates of the association between two species from
their conditional relationships (i.e. by controlling for other species in the
network) than we could get from their overall co-occurrence rates.

In this paper, following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I show that Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] can provide a framework
for understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observational data. Markov networks
have been used in many scientific fields for decades to model associations
between various kinds of "particles". For example, a well-studied network named the Ising model has played an
important role in our understanding of physics [where nearby particles tend to
align magnetically with one another; @cipra_introduction_1987]. In spatial
contexts, these models have been used to describe interactions between
adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
neurobiology, they have helped researchers determine which neurons are
connected to one another by modeling the structure in their firing patterns
[@schneidman_weak_2006]. While ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously imposed
severe approximations that produced unintelligible results [e.g.
"probabilities" greater than one; @gilpin_factors_1982].  Now that it is
computationally feasible to fit these models exactly, this kind of approach has become
worth a second look.

@@

\noindent  
***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
however, ecologists will often need to estimate these parameters from co-occurrence data.
When the number of species is reasonably small, one can compute exact maximum
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing
$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence
frequencies. For the analyses in this paper, I used the *rosalia* package
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
define the objective function and gradient as R code.  The rosalia package then
uses the `BFGS` method in R's `optim` function to find the best values for $\alpha$
and $\beta$.

@@

\noindent  
The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
a common null modeling technique reliably fails. Specifically, Markov networks
were successful even when direct interactions were largely overwhelmed by indirect
effects (Figure 1) or environmental effects (lower panels of Figure 3). For cases
where fitting a Markov network is computationally infeasible, these results also
indicate that partial covariances---which can be computed straightforwardly by linear
regression---can often provide a surprisingly useful approximation. The
partial correlations' success on simulated data may not
carry over to real data sets, however; @loh_structure_2013 show that the linear
approximations can be less reliable in cases where the true interaction matrix
contains more structure (e.g. guilds or trophic levels). On the other hand,
if ecologists are familiar enough with the natural history of their study
systems to describe this kind of structure as a prior distribution on the
parameters or as a penalty on the likelihood, then this information could
reduce the effective degres of freedom to estimate and real-world results might
be even better than those shown in Figure 3.

Ecologists will also need natural history to pin down the exact nature of the
interactions identified by a network model (e.g. which species in a
positively-associated pair is facilitating the other),  particularly when real
pairs of species can reciprocally influence one another in multiple ways
simultaneously [@bruno_inclusion_2003]; the $\beta$ coefficients in Markov
networks have to reduce this complexity  to a single number. In short, partial
correlations and Markov  networks both help prevent us from mistaking marginal
associations for conditional ones, but they can't tell us the underlying
biological mechanisms at work.

Despite these limitations, Markov networks have enormous potential to improve
ecological inferences. For example, Markov networks provide a simple answer
to the question of how competition should affect a species' overall
prevalence, which was a major flash point for the null model debates in the
1980’s [@roughgarden_competition_1983; @strong_ecological_1984]. Equation
1 can be used to calculate the expected prevalence of a species in the
absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012].
Competition's effect on prevalence in a Markov network can then be calculated
by subtracting this value and the observed prevalence (cf Figure 2D).

Markov networks---particularly the Ising model for binary networks---have been studied
in statistical physics for nearly a century [@cipra_introduction_1987],
and the models' properties,
capabilities, and limits are well-understood in a huge range of applications,
from spatial statistics [@gelfand_modelling_2005] to neuroscience
[@schneidman_weak_2006] to models of human behavior [@lee_statistical_2013].
Modeling species interactions using the same framework would thus allow
ecologists to tap into an enormous set of existing discoveries and techniques
for dealing with indirect effects, stability, and alternative stable states
(i.e. phase transitions; @cipra_introduction_1987).

This modeling approach is also highly
extensible, even when it is inconvenient to compute the likelihood exactly.
For example, the mistnet software package for joint species distribution modeling
[@harris_generating_2015] can fit *approximate* Markov networks to large
species assemblages (>300 species) while simultaneously modeling each species'
nonlinear response to the abiotic environment. Combining multiple ecological
processes into a common model could help ecologists to disentangle different
factors that can confound simpler co-occurrence analyses [cf @connor_checkered_2013].
Numerous other extensions are possible: Markov networks
can be fit with a mix of discrete and continuous variables, for example
[@lee_learning_2012]. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow the coefficient linking two species
in an interaction matrix to vary as a function of the abiotic environment or
of third-party species that tip the balance between facilitation and
exploitation [@bruno_inclusion_2003].

@@

\noindent  
***Acknowledgements:***
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, and N. J. Gotelli provided useful feedback on an earlier draft of
this work.

>>>>

While competition doesn't reliably leave the expected pattern in community
structure (a low co-occurrence rate at the landscape level), it nevertheless does
leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
there will be a deficit of co-occurrences between the competing shrub species,
and *among unshaded sites*, there will also be such a deficit. More generally,
we can obtain much better estimates of the association between two species from
their conditional relationships (i.e. by controlling for other species in the
network) than we could get from their overall co-occurrence rates. This kind of
control is difficult to obtain from null models, which begin with the assumptions
that all the pairwise interactions are zero and thus don’t need to be controlled for.

In this paper, following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I show that Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] can provide a framework
for understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observational data. Markov networks
have been used in many scientific fields for decades in similar contexts.
For example, a well-studied network named the Ising model has played an important
role in our understanding of physics [where nearby particles tend to align
magnetically with one another; @cipra_introduction_1987].
In spatial contexts, these models have been used to describe interactions between
adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
neurobiology, they have helped researchers identify neural connections from firing patterns
[@schneidman_weak_2006]. While ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously imposed
severe approximations that produced unintelligible results [e.g.
"probabilities" greater than one; @gilpin_factors_1982].  Now that it is
computationally feasible to fit these models exactly, this kind of approach is
worth a second look.

@@

\noindent  
***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
however, ecologists will often need to estimate these parameters from co-occurrence data.
When the number of species is reasonably small, one can compute exact maximum
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing
$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence
frequencies. For the analyses in this paper, I used the *rosalia* package
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
define the objective function and gradient as R code.  The rosalia package uses
standard convex optimization techniques to find the best values for $\alpha$ and $\beta$.

@@

\noindent  
The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
a common null modeling technique reliably fails. Specifically, Markov networks
were successful even when direct interactions were largely overwhelmed by indirect
effects (Figure 1). For cases where fitting a Markov network is computationally infeasible,
these results also indicate that partial covariances and generalized linear models (the two
methods that estimated conditional relationships rather than marginal ones) can both
provide useful approximations. The partial correlations' success on simulated
data may not carry over to real data sets, however; @loh_structure_2013 show that the linear
approximations can be less reliable in cases where the true interaction matrix
contains more structure (e.g. guilds or trophic levels).

It’s important to note that none of these methods can identify the exact nature of the
pairwise interactions (e.g. which species in a positively-associated pair is facilitating the other),
particularly when real pairs of species can reciprocally influence one another in multiple ways
simultaneously [@bruno_inclusion_2003]; with compositional data, there is only enough
information to provide a single number describing each species pair.  On the other hand, if
ecologists are familiar enough with the natural history of their study
systems that they can augment the likelihood function with a prior distribution or a suitable
penalty, then this information could reduce the effective degrees of freedom to estimate and
real-world results might be even better than those shown in Figure 3.

Despite these limitations, Markov networks have enormous potential to improve
ecological inferences. For example, Markov networks provide a simple answer
to the question of how competition should affect a species' overall
prevalence, which was a major flashpoint for the null model debates in the
1980’s [@roughgarden_competition_1983; @strong_ecological_1984]. Equation
1 can be used to calculate the expected prevalence of a species in the
absence of biotic influences [$\frac{1}{1 + e^{-\alpha}}$; @lee_learning_2012].
Competition's effect on prevalence in a Markov network can then be calculated
by subtracting this value and the observed prevalence (cf Figure 2D).  Apart from
BayesComm (which had the worst performance of any method evaluated here), none of
the other methods could produce this inference because they do not make predictions
about species composition. [[*back this up with more evidence or remove it*]]

Markov networks---particularly the Ising model for binary networks---have been studied
in statistical physics for nearly a century [@cipra_introduction_1987],
and the models' properties,
capabilities, and limits are well-understood in a huge range of applications.
Modeling species interactions using the same framework would thus allow
ecologists to tap into an enormous set of existing discoveries and techniques
for dealing with indirect effects, stability, and alternative stable states.
Numerous other extensions are possible: Markov networks
can be fit with a mix of discrete and continuous variables, for example
[@lee_learning_2012]. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow the coefficient linking two species
in an interaction matrix to vary as a function of the abiotic environment or
of third-party species that tip the balance between facilitation and
exploitation [@bruno_inclusion_2003].

@@

\noindent  
***Acknowledgements:***
This research was funded by a Graduate Research Fellowship from the US
National Science Foundation and benefited greatly from discussions with A.
Sih, M. L. Baskett, R. McElreath, R. J. Hijmans, A. C. Perry, and C. S. Tysor.
Additionally, A. K. Barner, E. Baldridge, E. P. White, D. Li, D. L. Miller, N.
Golding, N. J. Gotelli, and C. F. Dormann provided useful feedback on earlier drafts
of this work.
